---
title: "Package clinicaltrialsgov to download clinical data"
author: "Aleksandr V. Sokolov"
date: "2023-03-21"
output:
  md_document:
    variant: markdown_github
code_folding: show
---

## clinicaltrialsgov package

-   This package is designed to download data from ClinicalTrials.gov
-   It supports legacy and new APIs (as of 2022-2023)
-   Several search features are available
-   As of now only downloading full data is supported
-   The ClinicalTrials.gov beta (https://beta.clinicaltrials.gov/) is not yet implemented

```{r}
# Package installation
library(remotes)
remotes::install_github("AleksandrVSokolov/clinicaltrialsgov", force = TRUE)
```

We have installed the package

```{r}
# Importing
library(clinicaltrialsgov)
```

# Downloading trials with the new API

ClinicalTrials.gov provides convenient API to access its data. Derailed information could be read here: <https://clinicaltrials.gov/api/gui> As of March, 2023 only 100 studies could be downloaded per request with this API. The clinicaltrialsgov package currently supports downloading all information for trials

-   In the following code we download studies for lung cancer with API-acceptable expression

```{r}
# In this example we downloaded 50 trials for lung cancer (NOTE: max_rnk - min_rnk should be no more than 100)
dataset_1 <- clinicaltrialsgov::get_trials_API_query(query = "lung+cancer",
                                                    download_folder = "test",
                                                    min_rnk = 1,
                                                    max_rnk = 50)

```

This function is also creates a folder "test" and dumps there a corresponding JSON file with trials before concatenation

```{r}
list.files("test")
```


Preview of the data (we just show first 10 columns)

```{r}
head(dataset_1[,1:10])
```

Data may have a very big number of columns since every term obtained from a trial is stored in the independemt cell in the initial XML document

In some cases one might be willing to download more than 100 trials. In this case, it is necessary to perform to functions: clinicaltrialsgov::prepare_download_URLs() clinicaltrialsgov::get_trials_API_urls()

In the example below we will download \~ 400 trials for a drug olaparib

```{r}
urls <- clinicaltrialsgov::prepare_download_URLs(query = "olaparib")
writeLines(urls)
```

Now obtained URLs could be used to download trials

```{r}
dataset_2 <- clinicaltrialsgov::get_trials_API_urls(url_list = urls, download_folder = "test2")
```

Preview of the data (we just show first 10 columns)

```{r}
head(dataset_2[,1:10])
```

```{r}
nrow(dataset_2[,1:10])
str(dataset_2[,1:10])
```

# Downloading trials with legacy API

Legacy ClinicalTrials.gov API was based on the URL: "https://clinicaltrials.gov/ct2/results/download_fields?" and parameters from the advanced search "https://clinicaltrials.gov/ct2/search/advanced?"

- This API allows downloading very large sets of trials in the format as it is seen on the website
- clinicaltrialsgov package currently supports specifying conditions and treatments as well as additional parameters via URL "suffix"
- By default it downloads Interventional clinical trials

In the code below, we could see downloading trials for all combinations of several types of cancers with several treatments

```{r}

dataset_3 <- clinicaltrialsgov::clinical_trial_downloader_two_terms(
  condition_terms = c("brain cancer", "glioma"),
  treatment_terms = c("cisplatin", "temozolomide", "olaparib"),
  folder = "test_downloads",
  url_suffix = "&cntry=GB&state=&city=London")
```
Preview data (first 10 columns)
```{r}
nrow(dataset_3)
```

```{r}
head(dataset_3[,1:10])
```
In the folder test_downloads we could find individual download files and reporting file

```{r}
list.files("test_downloads")
```
Let's see the reporting file
```{r}
report_1 <- read.csv("test_downloads/reporting_file.csv")
report_1
```
All combinations of diseases and treatments are searched
```{r}
report_1[,c("Intervention", "Condition_searched")]
```

This function also supports specification of just disease or just therapy

```{r}
dataset_4 <- clinicaltrialsgov::clinical_trial_downloader_two_terms(
  treatment_terms = c("cisplatin", "temozolomide", "olaparib"),
  folder = "test_downloads2")
```
```{r}
nrow(dataset_4)
```




