---
title: "Package clinicaltrialsgov to download clinical data"
author: "Aleksandr V. Sokolov"
date: "2024-02-09"
output:
  md_document:
    variant: markdown_github
code_folding: show
---

## clinicaltrialsgov package

-   This package is designed to download data from ClinicalTrials.gov
-   It supports legacy and new APIs (as of 2021-2024)
-   Several search features are available, including specification of all search terms in advanced search
-   As of now only downloading full data is supported
-   The V2 version (2024) introduced support for ClinicalTrials.gov API, version 2.0 (https://clinicaltrials.gov/data-api/api)

```{r}
# Package installation
library(remotes)
remotes::install_github("AleksandrVSokolov/clinicaltrialsgov", force = TRUE)
```

We have installed the package

```{r}
# Importing
library(clinicaltrialsgov)
```

# Downloading clinical trials with the API version 2.0

This api provides support for both .CSV downloads (as in 2021) and as beta version developed in 2023. 
The process of downloading is simple and could be performed in few lines of code.

```{r}
# In this example we download all clinical trials with breast cancer, glioma, lung cancer that were treated with cisplatin or temozolomide or olaparib or dasatinib
clinical_trial_downloaderV2_API(condition_terms = c("breast cancer", "glioma", "lung cancer"),
                               treatment_terms = c("cisplatin", "temozolomide", "olaparib", "dasatinib"),
                               folder = "test_downloads_v2",
                               url_suffix = "&query.patient=AREA[LocationCity] London")
```

Note: we download clinical trials for all combinations between treatments and conditions. We specified via suffix "&query.patient=AREA[LocationCity] London" that we want to get trials specifically from London. Other text fields could be seen here: https://clinicaltrials.gov/data-api/api. To import clinical trials we can simply use the function import_ct_results_v2_csv() and specify a downloaded folder.

```{r}
results = import_ct_results_v2_csv("test_downloads_v2")
str(results)
```

Note: one of the combinations did not yield any result. To see it, simply read the file ZERO_URLS.txt located in the download directory

```{r}
readLines("test_downloads_v2/ZERO_URLS.txt")
```
It means that there was no clinical trial in London that was investigating glioma with a treatment of olaparib


# Downloading trials with the API beta-version 2023 (probably will expire soon, NOT RECOMMENDED)

Derailed information could be read here: <https://clinicaltrials.gov/api/gui> As of March, 2023 only 100 studies could be downloaded per request with this API. The clinicaltrialsgov package currently supports downloading all information for trials

-   In the following code we download studies for lung cancer with API-acceptable expression

```{r}
# In this example we downloaded 50 trials for lung cancer (NOTE: max_rnk - min_rnk should be no more than 100)
dataset_1 <- clinicaltrialsgov::get_trials_API_query(query = "lung+cancer",
                                                    download_folder = "test",
                                                    min_rnk = 1,
                                                    max_rnk = 50)

```

This function is also creates a folder "test" and dumps there a corresponding JSON file with trials before concatenation

```{r}
list.files("test")
```


Preview of the data (we just show first 10 columns)

```{r}
head(dataset_1[,1:10])
```

Data may have a very big number of columns since every term obtained from a trial is stored in the independemt cell in the initial XML document

In some cases one might be willing to download more than 100 trials. In this case, it is necessary to perform to functions: clinicaltrialsgov::prepare_download_URLs() clinicaltrialsgov::get_trials_API_urls()

In the example below we will download \~ 400 trials for a drug olaparib

```{r}
urls <- clinicaltrialsgov::prepare_download_URLs(query = "olaparib")
writeLines(urls)
```

Now obtained URLs could be used to download trials

```{r}
dataset_2 <- clinicaltrialsgov::get_trials_API_urls(url_list = urls, download_folder = "test2")
```

Preview of the data (we just show first 10 columns)

```{r}
head(dataset_2[,1:10])
```

```{r}
nrow(dataset_2[,1:10])
str(dataset_2[,1:10])
```

# Downloading trials with legacy API (version 2021, NOT RECOMMENDED)

Legacy ClinicalTrials.gov API was based on the URL: "https://clinicaltrials.gov/ct2/results/download_fields?" and parameters from the advanced search "https://clinicaltrials.gov/ct2/search/advanced?"

- This API allows downloading very large sets of trials in the format as it is seen on the website
- clinicaltrialsgov package currently supports specifying conditions and treatments as well as additional parameters via URL "suffix"
- By default it downloads Interventional clinical trials

In the code below, we could see downloading trials for all combinations of several types of cancers with several treatments

```{r}

dataset_3 <- clinicaltrialsgov::clinical_trial_downloader_two_terms(
  condition_terms = c("lung cancer", "glioma"),
  treatment_terms = c("cisplatin", "temozolomide", "doxorubicin"),
  folder = "test_downloads_old_api",
  url_suffix = "&cntry=GB&state=&city=London")
```
Preview data (first 10 columns)
```{r}
nrow(dataset_3)
```

```{r}
head(dataset_3[,1:10])
```
In the folder test_downloads we could find individual download files and reporting file

```{r}
list.files("test_downloads_old_api")
```
Let's see the reporting file
```{r}
report_1 <- read.csv("test_downloads_old_api/reporting_file.csv")
report_1
```
All combinations of diseases and treatments are searched
```{r}
report_1[,c("Intervention", "Condition_searched")]
```

This function also supports specification of just disease or just therapy

```{r}
dataset_4 <- clinicaltrialsgov::clinical_trial_downloader_two_terms(
  treatment_terms = c("cisplatin", "temozolomide", "olaparib"),
  folder = "test_downloads_old_api2")
nrow(dataset_4)
```



